\documentclass[a4paper]{article}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{algpseudocode} 
\usepackage{float}
\usepackage{booktabs}
\usepackage{amsmath}
\newtheorem{theorem}{Theorem}
%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{qtree}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{tikz}
\usetikzlibrary{automata,positioning}

\usepackage{forest}
\renewcommand{\rmdefault}{ptm}

\begin{document}
\section*{Karp-Rabin fingerprinting on strings}
Given a string $S = S[0 \dots n − 1]$, and two positions $0 \leq i < j \leq n − 1$, the longest common extension $lce_S(i, j)$ is the length of the maximal run of matching characters from those positions, namely: if $S[i] != S[j]$ then $lce_S(i, j) = 0$; otherwise, $lce_S(i, j) =  max \{ \textit{l}\geq 1 : S[i \dots i + \textit{l} − 1] = S[j \dots j + \textit{l} − 1] \}$. For example, if $S = abracadabra$, then $lce_S(1, 2) = 0$, $lce_S(0, 3) = 1$, and $lce_S(0, 7) = 4$. Given $S$ in advance for preprocessing, build a data structure for $S$ based on the Karp-Rabin fingerprinting, in O(n log n) time, so that it supports subsequent online queries of the following two types:
\begin{itemize}
\item $lce_S(i, j)$: it computes the longest common extension at positions $i$ and $j$ in O(log n) time.
\item $equalS(i, j,\textit{l})$: it checks if $S[i \dots i + \textit{l} − 1] = S[j \dots j + \textit{l} − 1]$ in constant time.
\end{itemize}
Analyze the cost and the error probability. The space occupied by the data structure can be O(n log n) but it is possible to use O(n) space. [Note: in this exercise, a onetime preprocessing is performed, and then many online queries are to be answered on the fly.]\\
\\
\textbf{SOLUTION}
\\
Karp-Rabin hashing on strings (i.e. $str[n]$)for solving the longest common extension problem. We have the following steps:
\begin{enumerate}
\item Create a data structure holding the hashes, with hashes of previous characters being held as a prefix. We fix a sufficient big prime $p$ and we use the Karp-Rabin hash (i.e. for a string $k$ and a base $b$: $h(k) = (k[0]b^{L−1} + k[1]b^{L−2}+ \dot+ k [L − 1]b^0) \ mod\  p $) :
\begin{align*}
        H[0]&   = h(str[0]) \\
        H[1]&   = H[0]  p + h(str[1])\\
        H[2]&   = H[1]  p + h(str[2]) \ = \ h(str[0]) p^2 + h(str[1]) p^1 +h(str[2])p^0\\
        &\cdots  \\
        H[n-1]& = H[n-2]  p + h(str[n-1]) \ = \  h(str[0]) p^{n-1} +  \dots  +h(str[n-2]) p^1 + h(str[n-1])p^0\\ 
\end{align*}
Therefore the space used here is just O(n).
\item Firstly we care about equality. To compare equality of a substring of length $l$ at indexes $i, j$, we need to know the sub-hash of the string. Thus, we take hashes of $H[i]$ and $H[i+l]$ and we compare calculate the hash of the sub-string between $i$ and $l$:
\begin{align*}
H[Substring_i] = &H[i + l] - H[i] * p ^{(l - 1)}\\
=&h(str[0]) p^{i + l-1} +  \dots  +h(str[i + l-2]) p^1 + h(str[i + l-1])p^0\\
-& (h(str[0]) p^{i-1} +  \dots  +h(str[i-2]) p^1 + h(str[i-1])p^0)* p ^{(l - 1)}\\
=& h(str[i + l-1])p^0+h(str[i + l-2]) p^1\dots +h(str[0]) p^{i + l-1}  \\
-& (h(str[i-1])p^{(l - 1)} +h(str[i-2]) p^{l}\dots +h(str[0]) p^{i+l - 2}    )\\
=& h(str[i + l-1])p^0+h(str[i + l-2]) p^1\dots +h(str[i]) p^{i + l-1}  \\   
\end{align*}
We do the same procedure for the sub-string between $j$ and $l$ (i.e. $H[Substring_j]$), the we can simply compare the two hash. We introduce a base case (a sanity check) in the case $l=1$ and we have $str[i] \neq str[j]$. The cost of this operation i O(1) since we did just simple operation ($+ \ *$).\\
We choose a random prime number $p\in [2,\cdots,\tau]$ where $\tau > n$. Since the prime number in the interval $[2,\cdots,\tau]$ are approximately $ \frac{\tau}{ln(\tau)}$, and we have a collision when $Substring_i=Substring_j$ but $H[Substring_i]\neq H[Substring_j]$, thus when $c= Substring_i- Substring_j \ mod \ p=0$. We can conclude that $P_r[error] \leq \frac{\textsc{\#bad prime}}{\textsc{\#prime}}= \frac{n}{\frac{\tau}{ln(\tau)}}$ because there are at most $n$ distinct prime $p$ that divide $c$ (Chinese Theorem of residual). If we choose $\tau \approx n^{a+1} ln(n)$ then we have $P_r[error] \leq \frac{1}{n^a}$.
\item Finally to compute the the longest common extension $lce_S(i, j)$, we just do a binary search of the index $l$. The cost to do so is O(ln(n)) since the check of the equality is constant.
\end{enumerate}

\newpage \qquad \\
\textbf{SOLUTION in O(N LOG N) works for every kind of hash function}
\\
The proposed data structure that maintain a series of trees. At each level we encode power of two elements (THIS IS IMPOSSIBLE TO EXPLAIN... LOOK THE CODE). The space occupied by this data structure is $O(nlog(n))$ where n is the length of the input array.   
\begin{algorithmic}
\Function{CreateTree}{$S$,$n$,$p$}
\State $A \gets NEW \ Array[log_2(n)+1]$
\State $TEMP \gets NEW \ Array[n]$
\For{$i \ IN \ (0,n)$}  
\State $TEMP[i] \gets S[i] \ mod \ p$
\EndFor 
\State $A[0] \gets TEMP$
\For{$i \ IN \ (1,log_2(n))$} 
\State $TEMP \gets NEW \ Array[n-i]$ 
\For{$j \ IN \ (0,n-i)$}  
\State $TEMP[j] \gets A[i-1][j] + A[i-1][j+2^{i-1}] \ mod \ p$
\EndFor 
\State $A[i] \gets TEMP$
\EndFor 
\State \Return $A$
\EndFunction
\end{algorithmic}
Now to have $lce_S(i, j)$ we build the implement the following procedure, where $h=\lfloor log_2(n-j)\rfloor$
\begin{algorithmic}
\Function{LCE}{$i$,$j$,$h$}
\If {$A[h][i]== A[h][j]$}
    \State \Return $2^h$
\Else
    \If {$h \neq 0$}
        \State \Return $0$
    \Else
    	\State \Return LCE$(i, j,h-1) +$LCE$(i+2^{h-1},j+2^{h-1},h-1)$ 
    \EndIf
\EndIf
\EndFunction
\end{algorithmic}
Notice that, this procedure work in $O(log(n))$ since the array is length is at most log(n) and we are doing two recursive call with an array one unit smaller each time. Finally to obtain $equalS(i, j,\textit{l})$ in cost $O(1)$ we simply check whether $A[\lceil log_2(l) \rceil )][i]== A[\lceil log_2(l) \rceil ][j]$ is true. \\
Notice that all this algorithm work for arrays in which their length $n$ is a power of two. If we have an array that is not of the latter length, we are doing the following: create the same data structure as before but the part of the array that is not in the tree, that it's at most long $2^{i+1}-2^{i}-1$ where $i= \lfloor log_2(n)\rfloor$, is store in simple array. In this case whether we need to check if $A[\lfloor log_2(n-j)\rfloor][i]== A[\lfloor log_2(n-j)\rfloor][j]$, i.e. the longest possible string, we need also to check, manually, whether there is a matching in the array. Last but not least, the calculation of the error probability is exactly the same to the one analysed during the course.  
\end{document}