\documentclass[a4paper]{article}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{algpseudocode} 
\usepackage{float}
\usepackage{booktabs}
\usepackage{amsmath}
\newtheorem{theorem}{Theorem}
%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{qtree}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{tikz}
\usetikzlibrary{automata,positioning}

\usepackage{forest}
\renewcommand{\rmdefault}{ptm}

\begin{document}
\section*{Hashing sets}
Your company has a database $S \in U$ of keys. For this database, it uses a randomly chosen hash function $h$ from a universal family $H$ (as seen in class); it also keeps a bit vector $B_S$ of $m$ entries, initialized to zeroes, which are then set $B_S[h(k)] = 1$ for every $k \in S$ (note that collisions may happen). Unfortunately, the database $S$ has been lost, thus only $B_S$ and $h$ are known, and the rest is no more accessible. Now, given $k \in U$, how can you establish if $k$ was in $S$ or not? What is the probability of error? Under the hypothesis that $m \geq c|S|$ for some $c > 1$ (note: we do not know the actual values of $c$ and $|S|$) can you estimate the size $|S|$, i.e. the size of $S$, looking at just $h$ and $B_S$? What is the probability of error? Note that $S$ is no more accessible as it disappeared.\\ 
Optional: Another database $R$ has been found to be lost: it was using the same hash function $h$, and the bit vector $B_R$ defined analogously as above. Using $h$, $B_S$, and $B_R$, how can you establish if $k$ was in $S \cup R$ (union), $S \cap R$ (intersection), or $S \setminus R$ (difference)? What is the probability of error?
\\
\textbf{SOLUTION}
\\
\end{document}
